Deriving Multi-Headed Projective Dependency Parses \\ from Link Grammar Parses

### 2014-05-30 doc/LinksToDAG.tex
updated the ILP formulation in the paper. Its really large now and does not fit in one column. But its much more complete of a description.

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}  % allows better color names
\usepackage{todonotes}   % insert [disable] to disable all notes
% Note that these macros accept optional arguments such as 
% [size=\small,bordercolor=red].
\newcommand{\Note}[4][]{\todo[author=#2,color=#3,fancyline,#1]{#4}}
\newcommand{\noteJH}[2][]{\Note[#1]{MRG}{blue!40}{#2}}
\newcommand{\noteJE}[2][]{\Note[#1]{JE}{green!40}{#2}}   
\newcommand{\notewho}[3][]{\Note[#1]{#2}{orange!40}{#3}}  % extra arg with miscellaneous author
\newcommand{\NoteJH}[2][]{\noteJH[inline,#1]{#2}}
\newcommand{\NoteJE}[2][]{\noteJE[inline,#1]{#2}}
\newcommand{\Notewho}[3][]{\notewho[inline,#1]{#2}{#3}}  % extra arg with miscellaneous author

@@

\section{Integer Linear Programming}
We formulated an ILP to assign directionality to all of the links in a set of link grammar parses, while still mandating that the resulting dependency parses are connected DAGs, with the nodes reachable from the root.

In order to choose which direction the link labels are assigned, for every encountered label we generate two boolean label/direction variables correspending to left and right. Setting one of these variables to \textsc{true} allows only the specified direction for a label throughout all parses, and because we assign a directionality to every link type at least one of these variables must be set to true. Setting both variables would allow us to go in either direction.

Our ILP objective is to minimize the number of label/direction variables set to \textsc{true}, while still satisfying the acyclicity and reachability constraints. The encoding was written in the Zimpl little language \cite{Koch2004} and solved using the SCIP Optimization Suite\cite{achterberg2009scip}.

@@

\begin{figure}
  \small
  \begin{align}
    \textsc{dir} &= \{\textsc{left}, \textsc{right}\} &\\ 
    \textsc{l} &\in \textsc{labels} & \\
    \textsc{m} &= \left|{\textsc{links}}\right| &\\
    \textsc{count}_{\textsc{l}, \textsc{dir}} &= \sum \textsc{link}_{\textsc{l},\textsc{dir}} &\\
    \textsc{count}_{\textsc{l}, \textsc{dir}} &= \textsc{m} \cdot \textsc{allowed}_{\textsc{l},\textsc{dir}} + \textsc{slack}_{\textsc{l}} &\\
    \textsc{cost}_{\textsc{l}} &= \frac{100}{\textsc{count}_{\textsc{l},\textsc{left}} + \textsc{count}_{\textsc{l},\textsc{right}}} &
  \end{align}
  \begin{equation} % I seperated out this line from the rest because it would disrupt the line number on the right side of the figure.
    \text{minimize:} \sum_{\textsc{l}, \textsc{dir}} (\textsc{allowed}_{\textsc{l},\textsc{dir}}) + \textsc{cost}_{\textsc{l}} \cdot \sum_\textsc{l} \textsc{slack}_{\textsc{l}} 
  \end{equation}
  \begin{align}
    \text{subject to } &\forall_{i \in \textsc{sentence}}: \textsc{reachable}_{\textsc{root}, i} = \textsc{true} &\\
                       &\forall_{\substack{i,j \in \textsc{sentence} \\ i\neq j }}: \textsc{has\_cycle}_{i,j} = \textsc{false} &
  \end{align}
  \caption{\small The ILP formulation.}
\end{figure}

\subsection{Slack Hierarchy}

We introduced slack on the directional dependency variable tokens such that link types were allowed to deviate from the majority up to (1\% ?) of the time before the ILP would assign both directional dependency variables to true. This addressed noise in the link parser's label assignments, while still allowing for the possibility that both directions could still be assigned.

\todo{} We also introduced slack on the link types such that link types with the same coarse grained label would try and align the same way as the majority in the group, where the preceding capital letters of the link type denote the coarse grained label, while the subscript letters denote further information. This slack places a prior on rare or never-before seen link types to be assigned in the same way as other similar link types.

This slack hierarchy gave the ILP the flexibility to handle noise and novel link types while still trying to learn the overall link grammar.

\subsection{Slack}
We introduced slack on the directional dependency variables such that the link types were allowed to deviate from the majority one percent of the time before the ILP would assign both directional dependency variables to true. This addressed noise in the link parser's label assignments, while still allowing for both directions to be assigned if needed.

\subsection{Stability of Results}

We report on the convergence of our ILP method to a stable annotation scheme as the corpus size increases.

We measured how the solution changed over increasing numbers of processed sentences. 
Taking the solution to the largest run as the answer key, we measured how much the previous runs deviated from it. We measured the precision of whether the assignments in the smaller runs matched the assignments of the largest run and we measured the recall of whether the assignments in the largest run could be found in the smaller runs. 
\begin{figure}[ht!]
  \small
  \includegraphics[width=\linewidth, keepaspectratio=true]{figure/precision_recall.png}
  \caption{\small Precision/recall of results of English sentences}
\end{figure}
We found that the precision values converged to 100\% while the recall values continued to grow throughout our data. This indicates that the solutions of the previous runs were consistent with the largest run, and as the data grew we would continue to encounter novel link types and incorporate them.

%\section{Other languages}

\section{Link Corpus Experiment and Results}

We will accomplish this by generating link parses using the AbiWord/CMU link grammar parser version 5.0.0 \noteJE{is this the version you used?} \cite{LINKPARSER-2014}, learning a direction to every encountered link label type, and then using these assignments for all link tokens in the link parse corpus. We formulate an integer linear programming problem that will find the smallest set of assignments that will still produce dependency parses that are acyclic with all nodes reachable from the root.

From the ILP, we produce a dictionary of label-to-direction assignments. We publish this so that anyone can easily turn a link parse into a dependency parse.

@@

\begin{figure}[ht!]
  \centering
  \small
  \input{figure/conll_analysis_sentences.tex}
  \caption{\small Of the sentences used for this experiment section, about \input{figure/conll_analysis_sentences_dropped}
    had at least one dropped word, and about \input{figure/conll_analysis_sentences_multiheaded}
    had at least one multiheaded word}
\end{figure}

@@

\begin{figure*}[ht!]
  \centering
  \input{figure/sentences.tikz}
  \caption{Example Sentences}
  \label{fig:parses}
\end{figure*}

@@

>>>>

\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}  % allows better color names
\usepackage{todonotes}   % insert [disable] to disable all notes
% Note that these macros accept optional arguments such as 
% [size=\small,bordercolor=red].
\newcommand{\Note}[4][]{\todo[author=#2,color=#3,fancyline,#1]{#4}}
\newcommand{\noteJH}[2][]{\Note[#1]{JH}{blue!40}{#2}}
\newcommand{\noteJE}[2][]{\Note[#1]{JE}{green!40}{#2}}   
\newcommand{\notewho}[3][]{\Note[#1]{#2}{orange!40}{#3}}  % extra arg with miscellaneous author
\newcommand{\NoteJH}[2][]{\noteJH[inline,#1]{#2}}
\newcommand{\NoteJE}[2][]{\noteJE[inline,#1]{#2}}
\newcommand{\Notewho}[3][]{\notewho[inline,#1]{#2}{#3}}  % extra arg with miscellaneous author

@@

\section{Integer Linear Programming}
We formulate an ILP to find the minimal set of link type to directionality assignments that would directionalize all the links in a set of link grammar parses, while still mandating that all of the resulting dependency parses would be connected DAGs, and with all the nodes reachable from the root.

For every encountered label we generate two boolean label/direction variables (e.g. label/left, label/right). Setting one of these variables to \textsc{true} allows the label type to only go in the specified direction, while setting both would allow the label to go in either way.
The objective is to minimize the number of label/direction variables set to \textsc{true}, while still satisfying acyclicity and reachability constraints. The encoding was written in the Zimpl little language \cite{Koch2004} and solved using the SCIP Optimization Suite\cite{achterberg2009scip}.

@@

\begin{figure*}
  \small
  \begin{align*}
    \text{Sets:}&&\\
    &\textsc{link}_{i,j,\textsc{label},\textsc{sentence}} \in \textsc{links} &\\
    &\textsc{label} \in \textsc{labels}&\\ 
    &\textsc{dir} = \{\textsc{left}, \textsc{right}\} &\\
    \text{Params:}&&\\
    &\textsc{m}_\textsc{length} = \max (\forall{\textsc{link}_{i,j,\textsc{sentence}}} \max(i,j)) &\\
    %&\textsc{count}_{\textsc{label}, \textsc{dir}} = \sum_{\textsc{link}_{\textsc{label},\textsc{dir}}}1 &\\
    &\textsc{cost}_{\textsc{label}} = \frac{100}{ \sum_{\textsc{link}_{\textsc{label},\textsc{dir}}}1 } &\\
    \text{Variables:}&&\\
    &\textsc{slack}_{\textsc{link}} \geq 0 &\\ 
    &\textsc{depth}_{i,\textsc{sentence}} \geq 0 &\\
    &\textsc{allowed}_{\textsc{label},\textsc{dir}} = \{0,1\}&\\
    &\textsc{arc}_{\textsc{link}} = \{0,1\} &\\
    \text{minimize:}& &\\ 
    &\sum_{\textsc{label}, \textsc{dir}} (\textsc{allowed}_{\textsc{label},\textsc{dir}}) + \textsc{cost}_{\textsc{label}} \cdot \sum_\textsc{link} \textsc{slack}_{\textsc{link}} &\\
    &&\\
    \text{subject to: }& &\\
    &\text{Links go in the allowed directions, except for slack}&\\
    &\forall{\textsc{link}_{i,j,\textsc{label}}}: 1-\textsc{arc}_{\textsc{link}} \leq \textsc{allowed}_{\textsc{label}, \textsc{left}} + \textsc{slack}_{\textsc{link}} &\\
    &\forall{\textsc{link}_{i,j,\textsc{label}}}: \textsc{arc}_{\textsc{link}} \leq \textsc{allowed}_{\textsc{label}, \textsc{right}} + \textsc{slack}_{\textsc{link}} &\\
    &\text{Depth constraints for acyclicity}&\\
    &\forall{\textsc{link}_{i,\textsc{sentence}}}: \textsc{depth}_{\text{root},\textsc{sentence}} \leq \textsc{depth}_{i,\textsc{sentence}} &\\
    &\forall{\textsc{link}_{i,j,\textsc{sentence}}}: \textsc{depth}_{i,\textsc{sentence}} + \textsc{m}_\textsc{length}\cdot\textsc{arc}_{i,j,\textsc{sentence}} \geq \textsc{depth}_{j,\textsc{sentence}} + 1 &\\
    &\forall{\textsc{link}_{i,j,\textsc{sentence}}}: \textsc{depth}_{j,\textsc{sentence}} + \textsc{m}_\textsc{length}\cdot(1-\textsc{arc}_{i,j,\textsc{sentence}}) \geq \textsc{depth}_{i,\textsc{sentence}} + 1 &\\
    &\text{Every token has a parent}&\\
    &\forall{\textsc{link}_{i \neq \textsc{root},\textsc{sentence}}}: \sum_{\textsc{link}_{i,j,\textsc{sentence}}} (1-\textsc{arc}_{i,j,\textsc{sentence}}) + \sum_{\textsc{link}_{j,i,\textsc{sentence}}}(\textsc{arc}_{j,i,\textsc{sentence}}) \geq 1&\\
  \end{align*}
  \caption{\small The ILP formulation.}
\end{figure*}

\subsection{Slack}
We introduced slack such that the link types were allowed to deviate from the majority one percent of the time before the ILP would assign both label/direction variables to \textsc{true}. This allows for flexibility against noise in the link parser's label assignments, balanced with allowing for both directions to be assigned if necessary.

\section{Experiments and Results}
To evaluate ...

We generated link parses using the AbiWord/CMU link grammar parser version 5.0.8 \cite{LINKPARSER-2014}.

The experiments were conducted on Russian and English sentences.
The Russian sentences were taken from a part of the Russian National Corpus (about 75,000 sentences)\noteJH{cite?}. 
The English sentences were taken from the conll bnews training data (about 60,000 sentences)\noteJH{cite?}. We were able to compare the resulting English link dependency data with the original conll to analyze the differences.

We publish the final link type to direction assignments so that anyone can easily turn a link parse into a dependency parse.

We report on the convergence of the results of our ILP with increasing data. With increasing trials of sentences, we measure how similar the subsequent directional assignments would be with each other. Taking the largest run as the answer key, we compared how much the smaller runs deviated from it. 
We recorded the precision of whether the assignments in the smaller runs could be found in those of the largest run, and similarly the recall of whether the assignments in the largest run could be found in the smaller runs.

\subsection{Stability of Results}
\begin{figure}[ht!]
  \small
  \centering
  \includegraphics[width=0.9\linewidth, keepaspectratio=true]{figure/precision_recall.png}
  \caption{\small Precision/recall of increasing runs of English and Russian sentences}
\end{figure}

We found that the precision values converged to 100\% while the recall values continued to grow throughout the data. This indicates that the solutions of the previous runs were consistent with the largest run, and as the data increased we would continue to encounter novel link types and incorporate them.

\subsection{Link Corpus Analysis}

\begin{figure*}[ht!]
  \centering
  \input{figure/sentences.tikz}
  \caption{Example Parses: The top blue arcs represent the conll dependencies, and the bottom red arcs are the directionalized link grammar parses}
  \label{fig:parses}
\end{figure*}

@@

\begin{figure}[ht!]
  \centering
  \small
  \input{figure/conll_analysis_sentences.tex}
  \caption{\small From the original sentences, we discarded disconnected parses for the ILP. From these sentences, we only looked at the complete link parses and discarded sentences with dropped words for our analysis section. About \input{figure/conll_analysis_sentences_dropped} of the non-disconnected sentences contained dropped words.}
\end{figure}

\todo{Fix the caption in the conll analysis sentence chart}

About \input{figure/conll_analysis_sentences_multiheaded} of the sentences with complete parses had at least one multiheaded word

@@

@@

Here we display a set of link parses as further examples
